From Reversible Programs to \\ Univalent Universes and Back

### 2017-05-04 upi.tex
examples; notation needs to be unified one way or the ohter; either all math; all haskell; all agda

The early developments of reversible programming languages started with a
conventional programming language, e.g., extended $\lambda$-calculi, and either
(i) extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
(ii) imposed constraints on the control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
recognize that reversible programming languages require a fresh approach and
should be designed from first principles without the detour via conventional
irreversible languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.

@@

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean
circuits~\cite{James:2012:IE:2103656.2103667}. With the addition of recursive
types and trace operators~\cite{Hasegawa:1997:RCS:645893.671607}, the variant
of $\Pi$ becomes a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}. Before identifying a
``featherweight'' representative to use for the formal development in this
paper, we illustrate the main underlying ideas for the entire family of
languages using some examples.

\amr{EXAMPLES OF CIRCUITS}

Although writing circuits using the raw syntax for combinators is tedious, the
examples illustrate the programming language nature of this family of $\Pi$
languages. In other work, one can find a compiler from a conventional functional
language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
Haskell-like surface language~\cite{theseus} which can ease writing
circuits. All that reinforces the claim that this is a practical approach to
reversible programming.

\amr{EXAMPLES OF OPTIMIZATIONS}

>>>>

The early developments of reversible programming languages started with a
conventional programming language, e.g., an extended $\lambda$-calculus, and either
(i) extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
(ii) imposed constraints on the control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
recognize that reversible programming languages require a fresh approach and
should be designed from first principles without the detour via conventional
irreversible languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.

@@

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean circuits~\cite{James:2012:IE:2103656.2103667}
and the extension with recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}. In this paper we will
identify a ``featherweight'' representative to use for the formal
development. But first, we illustrate the flavor of the family of languages with
some examples written in a Haskell syntax.

First consider the following program:
\begin{verbatim}
addSub1 :: (Int, Int) :<=> (Int, Int)
addSub1 = CommuteTimes
      :.: (UnfoldN :*: Id)
      :.: Distribute
      :.: (Id :+: CommuteTimes)
      :.: Factor
      :.: (FoldN :*: Id)
\end{verbatim}
which takes a pair of natural numbers and processes them as follows depending on
whether the second number is 0 or not. If the second number is 0, the evaluation
proceeds as follows:
$(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
\mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
$(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
\mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
(\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
applications of this program would increment the first number and decrement the
second number maintaining the sum at all times. When the second number reaches
0, the numbers are swapped.

As another example, consider the following expressions culminating with the
implementation of the Toffoli gate, whichi is universal for combinational
circuits.

\begin{verbatim}
inot :: Bool :<=> Bool
inot = UnfoldB :.: CommutePlus :.: FoldB

cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
cond f g = (UnfoldB :*: Id)
       :.: Distribute
       :.: ((Id :*: f) :+: (Id :*: g))
       :.: Factor
       :.: (FoldB :*: Id)

controlled :: (a :<=> a) -> ((Bool, a) :<=> (Bool, a))
controlled f = cond f Id

cnot :: (Bool, Bool) :<=> (Bool, Bool)
cnot = controlled inot

toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
\end{verbatim}

Although writing circuits using the raw syntax for combinators above is tedious,
the examples illustrate the programming language nature of this family of $\Pi$
languages. In other work, one can find a compiler from a conventional functional
language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
Haskell-like surface language~\cite{theseus} which can ease writing
programs. All that reinforces the claim that this is a practical approach to
reversible programming.

\begin{verbatim}
neg₁ neg₂ neg₃ neg₄ neg₅ : BOOL ⟷ BOOL
neg₁ = swap₊
neg₂ = id⟷ ◎ swap₊
neg₃ = swap₊ ◎ swap₊ ◎ swap₊
neg₄ = swap₊ ◎ id⟷
neg₅ = uniti⋆l ◎ swap⋆ ◎ (swap₊ ⊗ id⟷) ◎ swap⋆ ◎ unite⋆l
neg₆ = uniti⋆r ◎ (swap₊ {ONE} {ONE} ⊗ id⟷) ◎ unite⋆r

negEx : neg₅ ⇔ neg₁
negEx = uniti⋆l ◎ (swap⋆ ◎ ((swap₊ ⊗ id⟷) ◎ (swap⋆ ◎ unite⋆l)))
          ⇔⟨ id⇔ ⊡ assoc◎l ⟩
        uniti⋆l ◎ ((swap⋆ ◎ (swap₊ ⊗ id⟷)) ◎ (swap⋆ ◎ unite⋆l))
          ⇔⟨ id⇔ ⊡ (swapl⋆⇔ ⊡ id⇔) ⟩
        uniti⋆l ◎ (((id⟷ ⊗ swap₊) ◎ swap⋆) ◎ (swap⋆ ◎ unite⋆l))
          ⇔⟨ id⇔ ⊡ assoc◎r ⟩
        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ (swap⋆ ◎ (swap⋆ ◎ unite⋆l)))
          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ assoc◎l) ⟩
        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ ((swap⋆ ◎ swap⋆) ◎ unite⋆l))
          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ (linv◎l ⊡ id⇔)) ⟩
        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ (id⟷ ◎ unite⋆l))
          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ idl◎l) ⟩
        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ unite⋆l)
          ⇔⟨ assoc◎l ⟩
        (uniti⋆l ◎ (id⟷ ⊗ swap₊)) ◎ unite⋆l
          ⇔⟨ unitil⋆⇔l ⊡ id⇔ ⟩
        (swap₊ ◎ uniti⋆l) ◎ unite⋆l
          ⇔⟨ assoc◎r ⟩
        swap₊ ◎ (uniti⋆l ◎ unite⋆l)
          ⇔⟨ id⇔ ⊡ linv◎l ⟩
        swap₊ ◎ id⟷
          ⇔⟨ idr◎l ⟩
        swap₊ ▤
\end{verbatim}
